---
title: "Stats 198 Project"
subtitle: "Due May 2"
author: "Kina Desai, Allister Wolff, Florine Lijesen, Alexa Troob"
format: pdf
---

## Use of AI tools

Used for help w formatting for quadratic model



## Introduction 

Research question: Do higher income countries (measured by GDP per capita) have greater access to clean cooking fuels?
- potential to go further: 
Is access to clean cooking fuels associated with indoor air pollution death rates?


## Methodology

The methodology we will be using to investigate the research question of whether higher income countries (measured by GDP per capital) have greater access to clean cooking fuels is data from 2019 from the "Access to clean fuels for cooking vs. GDP per capita" dataset. The variable y (the outcome) is the percentage of the population in country i with access to clean cooking fuels and the variable x is the GDP per capita for country i. GDP per capita is skewed right, so to help analyze this data we have decided to use a log function to help linearize the relationship between x and y. 

The first model we use is a linear regression model. A linear regression model is used as opposed to a scatterpolot because if we were to plot GDP per capita versus access to clean cooking fuels, the data would be nonlinear. A logarithmic transformation of GDP allows the data to have linearity and thus is easier to interpret. Additionally, the linear regression model directly addresses the research question regarding the relationship between increasing GDP levels and access to clean cooking fuels. The equation for this linear regression model is Yi=β0+β1*log(Xi). 28.17, or the slope term β1, represents the average change in access to clean cooking fuels given a one-unit increase in log GDP per capita, while adjusting for all other predictors. 

The second model we use is an extended form of a linear regression model that includes quadratics. Adding the quadratic term to the linear regression model allows for a potential curvature of the data points to be observed. Also, looking at the data points shows that there may be evidence of diminishing increases in access to clean cooking fuels at higher income levels, which could be shown through a quadratic linear regression model. Since the ANOVA test shows a significant reduction in the residual sum of squares, we can conclude that this quadratic linear regression model is a better fit to the data than the above/previous linear regression model. The p-value, which is 1.471e-07, is statistically significant at the alpha equals 0.05 significance level and since the p-value is very small there is sufficient evidence to assume that this quadratic linear regression model reduces residual variation and is the better fit model compared to the above linear regression model. 

Additionally, we must evaluate the assumptions for the models. The assumption of linearity for the first model is likely violated since the residuals are not symmetrical around the y=0 line on the residual plot, but for the second model (quadratic linear regression model) the assumption of linearity is better satisfied. The second assumption is independence and for both models this is satisfied as 

```{r}
library(tidyverse)
data <- read_csv("access-to-clean-fuels-for-cooking-vs-gdp-per-capita.csv")

data_2019 <- data |>
  filter(Year == 2019)

ggplot(data_2019, aes(x = `GDP per capita`,
                      y = `Access to clean fuels and technologies for cooking`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  labs(x = "GDP per capita",
       y = "Access to clean cooking fuels (%)",
       title = "Access to clean cooking fuels vs GDP per capita (2019)")

```

```{r}
# Add income group column
data <- data |>
  mutate(income_group = case_when(
    `GDP per capita` < 1000  ~ "Low income",
    `GDP per capita` < 12000 ~ "Middle income",
    TRUE                     ~ "High income"
  ))|>
  filter(`Access to clean fuels and technologies for cooking` <= 100)  # remove bad values
```

```{r}
model1 <- lm(`Access to clean fuels and technologies for cooking` ~ 
              log(`GDP per capita`), 
            data = data_2019)
summary(model1)
```
```{r}
plot(model1$fitted.values,
     resid(model1),
     xlab = "Fitted values",
     ylab = "Residuals")

abline(h = 0, col = "red")
```

```{r}
library(tidyverse)
data_2019 <- data_2019 |>
  mutate(log_gdp = log(`GDP per capita`))

ggplot(data_2019, aes(x = log_gdp,
                      y = `Access to clean fuels and technologies for cooking`)) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "lm",
              formula = y ~ x + I(x^2)) +
  labs(x = "Log GDP per capita",
       y = "Access to clean cooking fuels (%)")
```
```{r}
model2 <- lm(`Access to clean fuels and technologies for cooking` ~ log(`GDP per capita`) + I(log(`GDP per capita`)^2),
             data = data_2019)
summary(model2)
```
```{r}
plot(model2$fitted.values,
     resid(model2),
     xlab = "Fitted values",
     ylab = "Residuals")

abline(h = 0, col = "red")
```

```{r}
anova(model1, model2)
```
```{r}
library(readr)
share_of_rural_population_with_electricity_access_vs_share_of_total_population_with_electricity_access <- read_csv("share-of-rural-population-with-electricity-access-vs-share-of-total-population-with-electricity-access.csv")
```
```{r}
rural_vs_urban <- share_of_rural_population_with_electricity_access_vs_share_of_total_population_with_electricity_access
```

```{r}
rural_vs_urban_2019 <- rural_vs_urban |>
  filter(Year == 2019)
```
```{r}
rural_vs_urban_2019 <- as.data.frame(rural_vs_urban_2019)
```
```{r}
joined_data <- full_join(data_2019, rural_vs_urban_2019, by = "Entity")
```
```{r}
joined_data <- joined_data |>
  mutate(rural_urban_gap = `Share of urban population with access` - 
          `Share of rural population with access`)
```
```{r}
model <- lm (`Access to clean fuels and technologies for cooking`~ log_gdp + rural_urban_gap, data = joined_data)
summary(model)
```
```{r}
library(readr)
deaths_attributed_to_household_air_pollution_vs_deaths_attributed_to_ambient_air_pollution <- read_csv("deaths-attributed-to-household-air-pollution-vs-deaths-attributed-to-ambient-air-pollution.csv")
```
```{r}
deaths <- deaths_attributed_to_household_air_pollution_vs_deaths_attributed_to_ambient_air_pollution
```
```{r}
deaths <- deaths |>
mutate(
    indoor_death_rate = `Deaths attributed to household air pollution` / Population * 100000
  )
```
```{r}
merged <- full_join(data_2019, deaths, by = "Entity")
```
```{r}

ggplot(merged, aes(x = `Access to clean fuels and technologies for cooking`,
                      y = `indoor_death_rate`)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Access to clean cooking fuels (%)",
       y = "Number of deaths from indoor air pollution/100,000 people",
       title = "Indoor air pollution death rate vs. access to clean cooking fuels (2019)")
```


```{r}
model3 <- lm(indoor_death_rate ~ `Access to clean fuels and technologies for cooking` + log_gdp,
  data = merged)
summary(model3)
```

```{r}
plot(model3$fitted.values,
     resid(model3),
     xlab = "Fitted values",
     ylab = "Residuals")

abline(h = 0, col = "red")
```

## Results 
1. When fitting a linear regression model, we fine that  GDP per capita is significantly associated to the access to clean cooking fuels. Specifically, a one-unit increase in log GDP per capita is associated with an average increase of 28.17% in access to clean cooking fuel. This relationshiop is highly statistically significant as p<0.001 providing evidsence of a positive association between a countries weatlh and their access to clean energy. 

2. 
When looking at the relationship between GDP per capita and the acess to clean fuel, the scatterplot and residual plot support the fact that the linearity assumption may not be satisfied. This is supported by the redisuals appearing in a curved pattern, indicating a nonlinear relationship.To address this, we tried fitting a quadratic relationship between GDP and access to clean fuel, which appeared to be a better fit, suggesting that for low-income countries, rise in GDP is important for increase fuel access, but this plateaus at a higher GDP level. ANOVA test affirms that the quadratic is a better fit, as it compares the linear and quadratic model to procvide a better fit.   

3. To look into the other factors associated with acess to ceal cooking fuels we added a control for inequalites within each country, which was the gap between urban and rural population regarding access to clean cooking fuels. While controlling for this inequality gap, a one-unit increase in log GDP is associated with an average increase of 20.34% in access to clean cooking fuels. This suggests that GDP is a very imnpoetant predictor of accsess. When holding log GDP constant, a one-unit increase in the gap in access between rural and urban population in a country is associated with a 0.58% decrease in access to clean cooking fuels, suggesting that equality within a country, despite national wealth, is also important for access. 

4. Lastlty, in Model 3 we looked ar the relationship between acsess to clean fuel and indoor air pollution (number of deaths from indoor air pollution/100,000 people), while adjesting for GDP per capita. While controlling for GDP, a one percent increase in access to clean cooking fuels is associated with an average decrease of 0.18 deaths per 100,000 people, which is statistically significant (p = 0.03). This suggests that access to clean cooking fuel has a negative association with indoor air pollution death rates not just GDP alone. Despite this, the model explains that there is not much variablility in dealth rates and the reisdual plot suggests that there are possible outliers and uneven spread. This means that the association is statistically significant, but should be understood cautiously and not assume a causal relationship. 

## Discussion 