---
title: "Stats 198 Project"
subtitle: "Due May 2"
author: "Kina Desai, Allister Wolff, Florine Lijesen, Alexa Troob"
format: pdf
---

## Use of AI tools

Used for help w formatting for quadratic model



## Introduction 

Research question: Do higher income countries (measured by GDP per capita) have greater access to clean cooking fuels?
- potential to go further: 
Is access to clean cooking fuels associated with indoor air pollution death rates?


## Methodology
```{r}
library(tidyverse)
data <- read_csv("access-to-clean-fuels-for-cooking-vs-gdp-per-capita.csv")

data_2019 <- data |>
  filter(Year == 2019)

ggplot(data_2019, aes(x = `GDP per capita`,
                      y = `Access to clean fuels and technologies for cooking`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  labs(x = "GDP per capita",
       y = "Access to clean cooking fuels (%)",
       title = "Access to clean cooking fuels vs GDP per capita (2019)")

```

```{r}
# Add income group column
data <- data |>
  mutate(income_group = case_when(
    `GDP per capita` < 1000  ~ "Low income",
    `GDP per capita` < 12000 ~ "Middle income",
    TRUE                     ~ "High income"
  ))|>
  filter(`Access to clean fuels and technologies for cooking` <= 100)  # remove bad values
```

```{r}
model1 <- lm(`Access to clean fuels and technologies for cooking` ~ 
              log(`GDP per capita`), 
            data = data_2019)
summary(model1)
```
```{r}
plot(model1$fitted.values,
     resid(model1),
     xlab = "Fitted values",
     ylab = "Residuals")

abline(h = 0, col = "red")
```

```{r}
library(tidyverse)
data_2019 <- data_2019 |>
  mutate(log_gdp = log(`GDP per capita`))

ggplot(data_2019, aes(x = log_gdp,
                      y = `Access to clean fuels and technologies for cooking`)) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "lm",
              formula = y ~ x + I(x^2)) +
  labs(x = "Log GDP per capita",
       y = "Access to clean cooking fuels (%)")
```
```{r}
model2 <- lm(`Access to clean fuels and technologies for cooking` ~ log(`GDP per capita`) + I(log(`GDP per capita`)^2),
             data = data_2019)
summary(model2)
```
```{r}
plot(model2$fitted.values,
     resid(model2),
     xlab = "Fitted values",
     ylab = "Residuals")

abline(h = 0, col = "red")
```

```{r}
anova(model1, model2)
```
```{r}
library(readr)
share_of_rural_population_with_electricity_access_vs_share_of_total_population_with_electricity_access <- read_csv("share-of-rural-population-with-electricity-access-vs-share-of-total-population-with-electricity-access.csv")
```
```{r}
rural_vs_urban <- share_of_rural_population_with_electricity_access_vs_share_of_total_population_with_electricity_access
```

```{r}
rural_vs_urban_2019 <- rural_vs_urban |>
  filter(Year == 2019)
```
```{r}
rural_vs_urban_2019 <- as.data.frame(rural_vs_urban_2019)
```
```{r}
joined_data <- full_join(data_2019, rural_vs_urban_2019, by = "Entity")
```
```{r}
joined_data <- joined_data |>
  mutate(rural_urban_gap = `Share of urban population with access` - 
          `Share of rural population with access`)
```
```{r}
model <- lm (`Access to clean fuels and technologies for cooking`~ log_gdp + rural_urban_gap, data = joined_data)
summary(model)
```
```{r}
library(readr)
deaths_attributed_to_household_air_pollution_vs_deaths_attributed_to_ambient_air_pollution <- read_csv("deaths-attributed-to-household-air-pollution-vs-deaths-attributed-to-ambient-air-pollution.csv")
```
```{r}
deaths <- deaths_attributed_to_household_air_pollution_vs_deaths_attributed_to_ambient_air_pollution
```
```{r}
deaths <- deaths |>
mutate(
    indoor_death_rate = `Deaths attributed to household air pollution` / Population * 100000
  )
```
```{r}
merged <- full_join(data_2019, deaths, by = "Entity")
```
```{r}

ggplot(merged, aes(x = `Access to clean fuels and technologies for cooking`,
                      y = `indoor_death_rate`)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Access to clean cooking fuels (%)",
       y = "Number of deaths from indoor air pollution/100,000 people",
       title = "Indoor air pollution death rate vs. access to clean cooking fuels (2019)")
```


```{r}
model3 <- lm(indoor_death_rate ~ `Access to clean fuels and technologies for cooking` + log_gdp,
  data = merged)
summary(model3)
```

```{r}
plot(model3$fitted.values,
     resid(model3),
     xlab = "Fitted values",
     ylab = "Residuals")

abline(h = 0, col = "red")
```

## Results 
1. When fitting a linear regression model, we fine that  GDP per capita is significantly associated to the access to clean cooking fuels. Specifically, a one-unit increase in log GDP per capita is associated with an average increase of 28.17% in access to clean cooking fuel. This relationshiop is highly statistically significant as p<0.001 providing evidsence of a positive association between a countries weatlh and their access to clean energy. 

2. 
When looking at the relationship between GDP per capita and the acess to clean fuel, the scatterplot and residual plot support the fact that the linearity assumption may not be satisfied. This is supported by the redisuals appearing in a curved pattern, indicating a nonlinear relationship.To address this, we tried fitting a quadratic relationship between GDP and access to clean fuel, which appeared to be a better fit, suggesting that for low-income countries, rise in GDP is important for increase fuel access, but this plateaus at a higher GDP level. ANOVA test affirms that the quadratic is a better fit, as it compares the linear and quadratic model to procvide a better fit.   

3. To look into the other factors associated with acess to ceal cooking fuels we added a control for inequalites within each country, which was the gap between urban and rural population regarding access to clean cooking fuels. While controlling for this inequality gap, a one-unit increase in log GDP is associated with an average increase of 20.34% in access to clean cooking fuels. This suggests that GDP is a very imnpoetant predictor of accsess. When holding log GDP constant, a one-unit increase in the gap in access between rural and urban population in a country is associated with a 0.58% decrease in access to clean cooking fuels, suggesting that equality within a country, despite national wealth, is also important for access. 

4. Lastlty, in Model 3 we looked ar the relationship between acsess to clean fuel and indoor air pollution (number of deaths from indoor air pollution/100,000 people), while adjesting for GDP per capita. While controlling for GDP, a one percent increase in access to clean cooking fuels is associated with an average decrease of 0.18 deaths per 100,000 people, which is statistically significant (p = 0.03). This suggests that access to clean cooking fuel has a negative association with indoor air pollution death rates not just GDP alone. Despite this, the model explains that there is not much variablility in dealth rates and the reisdual plot suggests that there are possible outliers and uneven spread. This means that the association is statistically significant, but should be understood cautiously and not assume a causal relationship. 

## Discussion 