---
title: "Stats 198 Project"
subtitle: "Due May 2"
author: "Kina Desai, Allister Wolff, Florine Lijesen, Alexa Troob"
format: pdf
---

## Use of AI tools

Used for help w formatting for quadratic model. Grammarly was used for grammar. 



## Introduction 

How are GDP per capita and rural–urban infrastructure disparities associated with access to clean cooking fuels and technologies, and how does clean cooking access relate to indoor air pollution mortality across countries?

## Methodology

The methodology we will be using to investigate the research question of whether higher income countries (measured by GDP per capital) have greater access to clean cooking fuels is data from 2019 from the "Access to clean fuels for cooking vs. GDP per capita" dataset. The variable y (the outcome) is the percentage of the population in country i with access to clean cooking fuels and the variable x is the GDP per capita for country i. GDP per capita is skewed right, so to help analyze this data we have decided to use a log function to help linearize the relationship between x and y. 

The first model we use is a linear regression model. A linear regression model is used as opposed to a scatterpolot because if we were to plot GDP per capita versus access to clean cooking fuels, the data would be nonlinear. A logarithmic transformation of GDP allows the data to have linearity and thus is easier to interpret. Additionally, the linear regression model directly addresses the research question regarding the relationship between increasing GDP levels and access to clean cooking fuels. The equation for this linear regression model is Yi=β0+β1*log(Xi). 28.17, or the slope term β1, represents the average change in access to clean cooking fuels given a one-unit increase in log GDP per capita, while adjusting for all other predictors. 

The second model we use is an extended form of a linear regression model that includes quadratics. Adding the quadratic term to the linear regression model allows for a potential curvature of the data points to be observed. Also, looking at the data points shows that there may be evidence of diminishing increases in access to clean cooking fuels at higher income levels, which could be shown through a quadratic linear regression model. Since the ANOVA test shows a significant reduction in the residual sum of squares, we can conclude that this quadratic linear regression model is a better fit to the data than the above/previous linear regression model. The p-value, which is 1.471e-07, is statistically significant at the alpha equals 0.05 significance level and since the p-value is very small there is sufficient evidence to assume that this quadratic linear regression model reduces residual variation and is the better fit model compared to the above linear regression model. 

Additionally, we must evaluate the assumptions for the models. The assumption of linearity for the first model is likely violated since the residuals are not symmetrical around the y=0 line on the residual plot, but for the second model (quadratic linear regression model) the assumption of linearity is better satisfied. The second assumption is independence and for both models this is satisfied as 

```{r}
library(tidyverse)
data <- read_csv("access-to-clean-fuels-for-cooking-vs-gdp-per-capita.csv")

data_2019 <- data |>
  filter(Year == 2019)

ggplot(data_2019, aes(x = `GDP per capita`,
                      y = `Access to clean fuels and technologies for cooking`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  labs(x = "GDP per capita",
       y = "Access to clean cooking fuels (%)",
       title = "Access to clean cooking fuels vs GDP per capita (2019)")

```

```{r}
# Add income group column
data <- data |>
  mutate(income_group = case_when(
    `GDP per capita` < 1000  ~ "Low income",
    `GDP per capita` < 12000 ~ "Middle income",
    TRUE                     ~ "High income"
  ))|>
  filter(`Access to clean fuels and technologies for cooking` <= 100)  # remove bad values
```

```{r}
model1 <- lm(`Access to clean fuels and technologies for cooking` ~ 
              log(`GDP per capita`), 
            data = data_2019)
summary(model1)
```
```{r}
plot(model1$fitted.values,
     resid(model1),
     xlab = "Fitted values",
     ylab = "Residuals")

abline(h = 0, col = "red")
```

```{r}
library(tidyverse)
data_2019 <- data_2019 |>
  mutate(log_gdp = log(`GDP per capita`))

ggplot(data_2019, aes(x = log_gdp,
                      y = `Access to clean fuels and technologies for cooking`)) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "lm",
              formula = y ~ x + I(x^2)) +
  labs(x = "Log GDP per capita",
       y = "Access to clean cooking fuels (%)")
```
```{r}
model2 <- lm(`Access to clean fuels and technologies for cooking` ~ log(`GDP per capita`) + I(log(`GDP per capita`)^2),
             data = data_2019)
summary(model2)
```
```{r}
plot(model2$fitted.values,
     resid(model2),
     xlab = "Fitted values",
     ylab = "Residuals")

abline(h = 0, col = "red")
```

```{r}
anova(model1, model2)
```
```{r}
library(readr)
share_of_rural_population_with_electricity_access_vs_share_of_total_population_with_electricity_access <- read_csv("share-of-rural-population-with-electricity-access-vs-share-of-total-population-with-electricity-access.csv")
```
```{r}
rural_vs_urban <- share_of_rural_population_with_electricity_access_vs_share_of_total_population_with_electricity_access
```

```{r}
rural_vs_urban_2019 <- rural_vs_urban |>
  filter(Year == 2019)
```
```{r}
rural_vs_urban_2019 <- as.data.frame(rural_vs_urban_2019)
```
```{r}
joined_data <- full_join(data_2019, rural_vs_urban_2019, by = "Entity")
```
```{r}
joined_data <- joined_data |>
  mutate(rural_urban_gap = `Share of urban population with access` - 
          `Share of rural population with access`)
```
```{r}
model <- lm (`Access to clean fuels and technologies for cooking`~ log_gdp + rural_urban_gap, data = joined_data)
summary(model)
```
```{r}
library(readr)
deaths_attributed_to_household_air_pollution_vs_deaths_attributed_to_ambient_air_pollution <- read_csv("deaths-attributed-to-household-air-pollution-vs-deaths-attributed-to-ambient-air-pollution.csv")
```
```{r}
deaths <- deaths_attributed_to_household_air_pollution_vs_deaths_attributed_to_ambient_air_pollution
```
```{r}
deaths <- deaths |>
mutate(
    indoor_death_rate = `Deaths attributed to household air pollution` / Population * 100000
  )
```
```{r}
merged <- full_join(data_2019, deaths, by = "Entity")
```
```{r}

ggplot(merged, aes(x = `Access to clean fuels and technologies for cooking`,
                      y = `indoor_death_rate`)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Access to clean cooking fuels (%)",
       y = "Number of deaths from indoor air pollution/100,000 people",
       title = "Indoor air pollution death rate vs. access to clean cooking fuels (2019)")
```


```{r}
model3 <- lm(indoor_death_rate ~ `Access to clean fuels and technologies for cooking` + log_gdp,
  data = merged)
summary(model3)
```

```{r}
plot(model3$fitted.values,
     resid(model3),
     xlab = "Fitted values",
     ylab = "Residuals")

abline(h = 0, col = "red")
```

## Results 
1. When fitting a linear regression model, we fine that  GDP per capita is significantly associated to the access to clean cooking fuels. Specifically, a one-unit increase in log GDP per capita is associated with an average increase of 28.17% in access to clean cooking fuel. This relationshiop is highly statistically significant as p<0.001 providing evidsence of a positive association between a countries weatlh and their access to clean energy. 

2. 
When looking at the relationship between GDP per capita and the acess to clean fuel, the scatterplot and residual plot support the fact that the linearity assumption may not be satisfied. This is supported by the redisuals appearing in a curved pattern, indicating a nonlinear relationship.To address this, we tried fitting a quadratic relationship between GDP and access to clean fuel, which appeared to be a better fit, suggesting that for low-income countries, rise in GDP is important for increase fuel access, but this plateaus at a higher GDP level. ANOVA test affirms that the quadratic is a better fit, as it compares the linear and quadratic model to procvide a better fit.   

3. To look into the other factors associated with acess to ceal cooking fuels we added a control for inequalites within each country, which was the gap between urban and rural population regarding access to clean cooking fuels. While controlling for this inequality gap, a one-unit increase in log GDP is associated with an average increase of 20.34% in access to clean cooking fuels. This suggests that GDP is a very imnpoetant predictor of accsess. When holding log GDP constant, a one-unit increase in the gap in access between rural and urban population in a country is associated with a 0.58% decrease in access to clean cooking fuels, suggesting that equality within a country, despite national wealth, is also important for access. 

4. Lastlty, in Model 3 we looked ar the relationship between acsess to clean fuel and indoor air pollution (number of deaths from indoor air pollution/100,000 people), while adjesting for GDP per capita. While controlling for GDP, a one percent increase in access to clean cooking fuels is associated with an average decrease of 0.18 deaths per 100,000 people, which is statistically significant (p = 0.03). This suggests that access to clean cooking fuel has a negative association with indoor air pollution death rates not just GDP alone. Despite this, the model explains that there is not much variablility in dealth rates and the reisdual plot suggests that there are possible outliers and uneven spread. This means that the association is statistically significant, but should be understood cautiously and not assume a causal relationship. 

## Discussion 

The findings of this study broadly support the central research question that economic development and infrastructure disparities are associated with access to clean cooking fuels and technologies across countries. Across all models, GDP per capita was positively associated with clean cooking access, suggesting that economic development plays an important role in expanding household energy infrastructure. However, the results also indicate that economic growth alone does not fully explain differences in clean cooking access, as disparities between rural and urban communities were also strongly associated with lower overall access rates.

In the initial linear regression model, higher GDP per capita was associated with an increase in clean cooking access. However, the residual plots revealed a lack of linearity. This prompted the use of a quadratic regression model, which provided a significantly better fit. The quadratic results suggest that access improvements begin to plateau at higher income levels, a pattern consistent with diminishing returns. This is where economic growth may have a disproportionately larger impact in lower-income countries than in nations where there is near universal access.
  
Adding measures of rural and urban disparities further strengthened the analysis, showing that economic development alone does not fully explain disparities in clean cooking access. Even after controlling for GDP per capita, countries with wider disparities between rural and urban electrification tended to report lower overall access to clean cooking fuels. This finding suggests that access to clean cooking fuels is influenced not only by how wealthy a country is overall, but also by how evenly resources and infrastructure are distributed within the country. Even in countries with higher GDP per capita, rural communities and underserved populations may still face barriers to accessing reliable, clean energy sources. Economic growth alone may not guarantee equal improvements in living conditions for all populations.

The final model examining indoor air pollution mortality also supports the broader public health implications of clean cooking access. Countries with greater access to clean cooking fuels were found to experience lower indoor air pollution death rates. This aligns with existing research that has linked increased indoor air pollution to respiratory and cardiovascular disease. However, the model suggests that clean cooking access and GDP per capita explain only a small proportion of variation in indoor air pollution mortality due to the low adjusted R2 value. This likely means that additional factors such as healthcare access, housing conditions, environmental regulations, smoking prevalence, and ventilation quality also affect the variation in indoor air pollution mortality. Therefore, although the relationship observed was statistically significant, the results should be interpreted cautiously and not as evidence of direct causation.

It is important to note, however, that there are some limitations with this study. The first one is that GDP per capita was used as an indicator of wealth, and while it is a good way to measure average wealth, it can mask extreme cases of poverty and fails to account for cost-of-living differences, which influence access to clean cooking fuels. Further, in this data set, observations were excluded that were missing data which could have impacted the results. It is found that many high-income countries have better and more reliable reporting systems than low-income countries, which potentially biases the results. Another limitation is that the study relied on cross-sectional data that only spanned one year, which limits the ability to evaluate differences over time, reducing causality. The study can determine correlations between GDP and cooking levels; however, it does not have the evidence to make statements about whether increases in GDP directly cause improvements in clean cooking access. Lastly, these data sets were obtained from reputable sources; however, there still may be concerns about the reliability and validity of the data. This can be attributed to the fact that we do not know how they collected the data in each country and what each of their reporting standards is. A key part of this difference in collection could be, as mentioned earlier, that estimates of indoor air pollution mortality may be less accurate in low-income settings, as they have limited healthcare infrastructure and a less reliable reporting system. Additionally, measures such as “access to clean cooking fuels” may also be inaccurate to the actual situation, as some have access but cannot afford to use it for every meal, or fuel supplies are unreliable. 

If this study were to be performed again, we would use a longitudinal analysis that spans multiple years to strengthen the ability to examine trends and changes over time. Finding more data that includes variables such as education level, healthcare access, government energy policies, or income inequality could help strengthen the analysis, similar to what we saw with looking at urban and rural settings. Additionally, using data that is from the household level rather than country-level data could help account for the limitations mentioned. By accounting for the feasibility and fidelity of using clean cooking fuels, there will be more accurate data from low-resource settings. Looking into the future, additional research could investigate whether targeted policy interventions or infrastructure investments improve access to clean cooking fuels independently of broader economic growth. Government subsidization of clean cooking fuels or investments in rural gas infrastructure could potentially play a role in helping increase access to clean cooking fuels in low-income households. Addressing these gaps, such as socioeconomic status and living situation, will be essential to finding ways to increase access to clean cooking fuels.

Overall, the findings suggest that increasing access to clean cooking fuels is closely tied to both economic development and infrastructure equity. Expanding clean household energy access may not only improve quality of life but also reduce preventable health burdens associated with indoor air pollution, particularly in lower-income and rural populations.









