---
title: "Stats 198 Project"
subtitle: "Due May 2"
author: "Kina Desai, Allister Wolff, Florine Lijesen, Alexa Troob"
format: pdf
---

## Use of AI tools

Used for guidance with formatting and equation for quadratic model since we didn't directly learn this in class. Grammarly was used for grammar. 

## Introduction 

Access to clean cooking fuels and technologies remains an important global public health and development issue. Even as energy infrastructure has advanced, many households worldwide still rely on fuels such as wood, charcoal, coal, and agricultural waste for daily cooking. Burning these specific fuels indoors produces harmful air pollution that has been linked to serious health problems, including respiratory and cardiovascular disease. The World Health Organization has identified exposure to household air pollution as one of the leading causes of preventable deaths, especially in low and middle-income countries. As a result, improving access to clean cooking technologies has become an important target of global health and sustainability initiatives to reduce health disparities and improve quality of life.


Given the importance of this issue, this study aims to answer the question: are national wealth, measured by GDP per capita,  and within-country infrastructure disparities associated with access to clean cooking fuels and technologies across countries? In particular, we are examining if GDP per capita is positively associated with access to clean cooking fuels, and whether rural-urban inequalities within countries, measured by differences in access to electricity, explain additional variation in clean fuel access beyond GDP. We further investigate whether access to clean cooking fuels is associated with indoor air pollution mortality rates, while adjusting for GDP. All data used in this analysis was gathered from air pollution datasets compiled by Our World in Data, including indicators on fuel access, economic conditions, and health outcomes. 

Key variables used in our analysis include: 

Access to clean fuels and technologies for cooking = Percent of population with clean cooking fuel and technology access 

log_gdp = A country's logged GDP per capita 

rural_urban_gap = Share of urban population with access to clean cooking fuels minus the share of rural population with access to clean cooking fuels in a country. 

indoor_death_rate = Total number of deaths attributed to indoor air pollution in a country divided by the country's population and multiplied by 100,000 to get the mortality rate attributable to indoor air pollution per 100,000 people.  

## Methodology

To investigate whether higher income countries (measured by GDP per capita) are predicted to have greater access to clean cooking fuels, we configured models from the datatset "Access to clean fuels for cooking vs. GDP per capita" using the year 2019 since this was the most recent year with data available across all variables. The variable y (the outcome) is the percentage of the population in country i with access to clean cooking fuels and the variable x is the GDP per capita for country i. GDP per capita is skewed right, so to help analyze this data we have decided to use a log function to help linearize the relationship between x and y. 

The first model we use is a linear regression model. If we were to plot a scatterplot of GDP per capita versus access to clean cooking fuels, the data would be nonlinear. Therefore, we used a logarithmic transformation of GDP to allows the data to have linearity and be interpreted more easily. Additionally, the linear regression model directly addresses the research question regarding the relationship between increasing GDP levels and access to clean cooking fuels. We then assessed for assumptions.The assumption of independence is likely satisfied as knowing something about one country should not tell you anything about another country (even though this may be imperfect). Normality also appears to be satisfied since the histogram of the residuals follow an approximately normal distribution. However, we found that the assumption of linearity is likely violated since the residuals are not symmetrical around the y=0 line on the residual plot. The assumption of constant variance appears to be violated as well since the spread of residuals changes across fitted values. 

Therefore, the second model is also a linear regression model, but fitted with a quadratic relationship to better satisfy assumptions. Independence and normality are still satisfied under this second model, and the assumptions of linearity and constant variance are better satisfied using this model since the residuals are more symmetrical around the y=0 line and more evenly spread. Adding the quadratic term to the linear regression model allows for a potential curvature of the data points to be observed. This model suggests that there may be a diminishing increases in access to clean cooking fuels at higher income levels. Since the ANOVA test shows a significant reduction in the residual sum of squares, we can conclude that this quadratic linear regression model is a better fit to the data than the above/previous linear regression model that uses a linear relationship. The p-value, which is 1.471e-07, is statistically significant at the alpha equals 0.05 significance level so there is sufficient evidence to assume that this quadratic linear regression model reduces residual variation and is the better fit model.

This methodology focuses on estimating associations rather than causal relationships, given the observational and cross-sectional nature of the data. The linear regression model with a quadratic relationship was chosen as the preferred methodology because it is more easily interpreted, better captures nonlinear patterns, and is the better fit for the data points in the model. Furthermore, this methodology is easily reproducible by other researchers. To reproduce these results, one would filter the dataset to include only the year 2019, remove missing or invalid values, use the log GDP per capita, fit the linear regression model and quadratic linear regression model, and assess the assumptions. 

Our following linear regression model incorporated another predictor variable, the disparity between rural and urban access to electricity, as a control by joining the dataset, "Share or rural vs. urban population with electricity access," to our previously used dataset. Our final linear regression model used mortality rate from indoor air pollution as the outcome variable and access to clean fuels and GDP per capita as the predictor variables by joining a third dataset "Deaths attributed to household air pollution vs deaths attributed to ambient air pollution." We used the deaths attributed to household air pollution data and divided it by the country's population, then multiplied it by 100,000, to get the number of deaths due to household air pollution per 100,000 people in a country. 

## Model 1 : Linear Fit

```{r}
#| echo: false
#| message: false
library(tidyverse)
data <- read_csv("access-to-clean-fuels-for-cooking-vs-gdp-per-capita.csv")

data_2019 <- data |>
  filter(Year == 2019)

ggplot(data_2019, aes(x = `GDP per capita`,
                      y = `Access to clean fuels and technologies for cooking`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  labs(x = "GDP per capita",
       y = "Access to clean cooking fuels (%)",
       title = "Access to clean cooking fuels vs GDP per capita (2019)")

```

```{r}
#| echo: false
#| message: false
# Add income group column
data <- data |>
  mutate(income_group = case_when(
    `GDP per capita` < 1000  ~ "Low income",
    `GDP per capita` < 12000 ~ "Middle income",
    TRUE                     ~ "High income"
  ))|>
  filter(`Access to clean fuels and technologies for cooking` <= 100)  # remove bad values
```

```{r}
#| echo: false
#| message: false
model1 <- lm(`Access to clean fuels and technologies for cooking` ~ 
              log(`GDP per capita`), 
            data = data_2019)
summary(model1)
```
```{r}
#| echo: false
#| message: false
plot(model1$fitted.values,
     resid(model1),
     xlab = "Fitted values",
     ylab = "Residuals")

abline(h = 0, col = "red")
```
```{r}
#| echo: false
#| message: false
hist(resid(model1),
     main = "Histogram of Residuals (Model 1)",
     xlab = "Residuals",
     breaks = 20)
```
## Model 2: Quadratic fit 
```{r}
#| echo: false
#| message: false
library(tidyverse)
data_2019 <- data_2019 |>
  mutate(log_gdp = log(`GDP per capita`))

ggplot(data_2019, aes(x = log_gdp,
                      y = `Access to clean fuels and technologies for cooking`)) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "lm",
              formula = y ~ x + I(x^2)) +
  labs(x = "Log GDP per capita",
       y = "Access to clean cooking fuels (%)")
```
```{r}
#| echo: false
#| message: false
model2 <- lm(`Access to clean fuels and technologies for cooking` ~ log(`GDP per capita`) + I(log(`GDP per capita`)^2),
             data = data_2019)
summary(model2)
```
```{r}
#| echo: false
#| message: false
plot(model2$fitted.values,
     resid(model2),
     xlab = "Fitted values",
     ylab = "Residuals")

abline(h = 0, col = "red")
```
```{r}
#| echo: false
#| message: false
hist(resid(model2),
     main = "Histogram of Residuals (Model 2)",
     xlab = "Residuals",
     breaks = 20)
```


```{r}
#| echo: false
#| message: false
anova(model1, model2)
```
```{r}
#| echo: false
#| message: false
library(readr)
share_of_rural_population_with_electricity_access_vs_share_of_total_population_with_electricity_access <- read_csv("share-of-rural-population-with-electricity-access-vs-share-of-total-population-with-electricity-access.csv")
```
```{r}
#| echo: false
#| message: false
rural_vs_urban <- share_of_rural_population_with_electricity_access_vs_share_of_total_population_with_electricity_access
```

```{r}
#| echo: false
#| message: false
rural_vs_urban_2019 <- rural_vs_urban |>
  filter(Year == 2019)
```
```{r}
#| echo: false
#| message: false
rural_vs_urban_2019 <- as.data.frame(rural_vs_urban_2019)
```
```{r}
#| echo: false
#| message: false
joined_data <- full_join(data_2019, rural_vs_urban_2019, by = "Entity")
```
```{r}
#| echo: false
#| message: false
joined_data <- joined_data |>
  mutate(rural_urban_gap = `Share of urban population with access` - 
          `Share of rural population with access`)
```
## Adding a predictor variable: within-country access gap  
```{r}
#| echo: false
#| message: false
model <- lm (`Access to clean fuels and technologies for cooking`~ log_gdp + rural_urban_gap, data = joined_data)
summary(model)
```
```{r}
#| echo: false
#| message: false
library(readr)
deaths_attributed_to_household_air_pollution_vs_deaths_attributed_to_ambient_air_pollution <- read_csv("deaths-attributed-to-household-air-pollution-vs-deaths-attributed-to-ambient-air-pollution.csv")
```
```{r}
#| echo: false
#| message: false
deaths <- deaths_attributed_to_household_air_pollution_vs_deaths_attributed_to_ambient_air_pollution
```
```{r}
#| echo: false
#| message: false
deaths <- deaths |>
mutate(
    indoor_death_rate = `Deaths attributed to household air pollution` / Population * 100000
  )
```
```{r}
#| echo: false
#| message: false
merged <- full_join(data_2019, deaths, by = "Entity")
```
## Model 3: Access to clean fuels vs. indoor death rate  
```{r}
#| echo: false
#| message: false
ggplot(merged, aes(x = `Access to clean fuels and technologies for cooking`,
                      y = `indoor_death_rate`)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Access to clean cooking fuels (%)",
       y = "Number of deaths from indoor air pollution/100,000 people",
       title = "Indoor air pollution death rate vs. access to clean cooking fuels (2019)")
```

```{r}
#| echo: false
#| message: false
model3 <- lm(indoor_death_rate ~ `Access to clean fuels and technologies for cooking` + log_gdp,
  data = merged)
summary(model3)
```

```{r}
#| echo: false
#| message: false
plot(model3$fitted.values,
     resid(model3),
     xlab = "Fitted values",
     ylab = "Residuals")

abline(h = 0, col = "red")
```

## Results 

1. When fitting a linear regression model, we find that GDP per capita is significantly associated with access to clean cooking fuels. Specifically, a one-unit increase in log GDP per capita is associated with an average increase of 28.17 percentage points in access to clean cooking fuels. This relationship is highly statistically significant as p<0.001, providing evidence of a positive association between a country's wealth and their access to clean energy. 

2. While the above findings were statistically significant, the residual plot of the model suggests that the linearity and constant variance assumptions may not be satisfied. This is supported by the residuals appearing in a curved pattern, indicating a nonlinear relationship. To address this, we tried fitting a quadratic relationship between GDP and access to clean fuel, which appeared to be a better fit, suggesting that for low-income countries, rise in GDP is important for increase fuel access, but this plateaus at a higher GDP level when access becomes more universal. ANOVA test confirms that the quadratic provides a better fit than the linear relationship in model 1.  

3. To look into and adjust for other factors associated with access to clean cooking fuels, we added a control for disparities within each country, which we identified as the gap between urban and rural access to clean cooking fuels. When controlling for this within country rural-urban inequality, a one-unit increase in log GDP per capita is associated with an average increase of 20.34% in access to clean cooking fuels. This suggests that GDP is a very important predictor of access, regardless of unequal access within each country. Furthermore, when holding log-GDP constant, a one-unit increase in the gap in access between rural and urban populations in a country is associated with a 0.58% decrease in access to clean cooking fuels. This is statistically significant, with a p-value of 9.39e-12, suggesting that equality and access distribution within a country, despite national wealth, is also important for access to clean cooking fuels. 

4. Lastly, in Model 3 we looked at the relationship between access to clean fuel and deaths from indoor air pollution (number of deaths from indoor air pollution/100,000 people), while adjusting for GDP per capita. By directly connecting access to clean fuel to mortality rates, we were hoping to gain a deeper understanding of the public health significance of our previous findings. While controlling for GDP, a one percent increase in access to clean cooking fuels is associated with an average decrease of 0.18 deaths per 100,000 people, which is statistically significant (p = 0.03). This suggests that access to clean cooking fuel has a negative association with indoor air pollution death rates, not just GDP alone. Despite this, the model explains that there is not much variability in death rates and the residual plot suggests that there are possible outliars and uneven spread. This means that the association is statistically significant, but should be understood cautiously and not assume a causal relationship. 

## Discussion 

The findings of this study broadly support the central research question that economic development and infrastructure disparities are associated with access to clean cooking fuels and technologies across countries. Across all models, GDP per capita was positively associated with clean cooking fuel access, suggesting that economic development plays an important role in expanding household energy infrastructure. However, the results also indicate that economic growth alone does not fully explain differences in clean cooking access, as disparities between rural and urban communities were also strongly associated with lower overall access rates.  

In the initial linear regression model, higher GDP per capita was associated with an increase in clean cooking access. However, the residual plots revealed a lack of linearity. This prompted the use of a quadratic regression model, which provided a significantly better fit. The quadratic results suggest that access improvements begin to plateau at higher income levels, a pattern consistent with diminishing returns. This is where economic growth may have a disproportionately larger impact in lower-income countries than in nations where there is near universal access. This suggests that economic growth has the highest returns for clean fuel access in lower-income countries. 
  
Adding measures of rural and urban disparities further strengthened the analysis, showing that economic development alone does not fully explain disparities in clean cooking access. Even after controlling for GDP per capita, countries with wider disparities between rural and urban electrification tended to report lower overall access to clean cooking fuels. This finding suggests that access to clean cooking fuels is influenced not only by how wealthy a country is overall, but also by how evenly resources and infrastructure are distributed within the country. Even in countries with higher GDP per capita, rural communities and underserved populations may still face barriers to accessing reliable, clean energy sources. Economic growth alone may not guarantee equal improvements in living conditions for all populations.

The final model examining indoor air pollution mortality also supports the broader public health implications of clean cooking access. Countries with greater access to clean cooking fuels were found to experience lower indoor air pollution death rates. This aligns with existing research that has linked increased indoor air pollution to respiratory and cardiovascular disease. However, the model suggests that clean cooking access and GDP per capita explain only a small proportion of variation in indoor air pollution mortality due to the low adjusted R2 value. This likely means that additional factors such as healthcare access, housing conditions, environmental regulations, smoking prevalence, and ventilation quality also affect the variation in indoor air pollution mortality. Therefore, although the relationship observed was statistically significant, the results should be interpreted cautiously and not as evidence of direct causation.

It is important to note, however, that there are some limitations with this study. The first one is that GDP per capita was used as an indicator of wealth, and while it is a good way to measure average wealth, it can mask extreme cases of poverty and fails to account for cost-of-living differences, which influence access to clean cooking fuels. Further, in this data set, observations were excluded that were missing data which could have impacted the results. Many high-income countries have better and more reliable reporting systems than low-income countries, which potentially biases the results. Another limitation is that the study relied on cross-sectional data that only spanned one year, which limits the ability to evaluate differences over time, reducing causality. The study can determine correlations between GDP and cooking levels; however, it does not have the evidence to make statements about whether increases in GDP directly cause improvements in clean cooking access. Lastly, these data sets were obtained from reputable sources; however, there still may be concerns about the reliability and validity of the data. This can be attributed to the fact that we do not know how they collected the data in each country and what each of their reporting standards is. A key part of this difference in collection could be, as mentioned earlier, that estimates of indoor air pollution mortality may be less accurate in low-income settings, as they have limited healthcare infrastructure and a less reliable reporting system. Additionally, measures such as “access to clean cooking fuels” may also be inaccurate to the actual situation, as some have access but cannot afford to use it for every meal, or fuel supplies are unreliable. 

If this study were to be performed again, we would use a longitudinal analysis that spans multiple years to strengthen the ability to examine trends and changes over time. Finding more data that includes variables such as education level, healthcare access, government energy policies, or income inequality could help strengthen the analysis, similar to what we saw with looking at urban and rural settings. Additionally, using data that is from the household level rather than country-level data could help account for the limitations mentioned. By accounting for the feasibility and fidelity of using clean cooking fuels, there will be more accurate data from low-resource settings. Looking into the future, additional research could investigate whether targeted policy interventions or infrastructure investments improve access to clean cooking fuels independently of broader economic growth. Government subsidization of clean cooking fuels or investments in rural gas infrastructure could potentially play a role in helping increase access to clean cooking fuels in low-income households. Addressing these gaps, such as socioeconomic status and living situation, will be essential to finding ways to increase access to clean cooking fuels.

Overall, the findings suggest that increasing access to clean cooking fuels is closely tied to both economic development and infrastructure equity. Expanding clean household energy access may not only improve quality of life but also reduce preventable health burdens associated with indoor air pollution, particularly in lower-income and rural populations.

