---
title: "Stats 198 Project"
subtitle: "Due Wednesday, April 22"
author: "Kina Desai, Allister Wolff, Florine Lijesen"
format: pdf
---

## Use of AI tools

Used for help w formatting for quadratic model



## Introduction 

Research question: Do higher income countries (measured by GDP per capita) have greater access to clean cooking fuels?
- potential to go further: 




## Methodology
```{r}
library(tidyverse)
data <- read_csv("access-to-clean-fuels-for-cooking-vs-gdp-per-capita.csv")

data_2023 <- data |>
  filter(Year == 2023)

ggplot(data_2023, aes(x = `GDP per capita`,
                      y = `Access to clean fuels and technologies for cooking`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  labs(x = "GDP per capita",
       y = "Access to clean cooking fuels (%)",
       title = "Access to clean cooking fuels vs GDP per capita (2023)")

```
```{r}
# Add income group column
data <- data |>
  mutate(income_group = case_when(
    `GDP per capita` < 1000  ~ "Low income",
    `GDP per capita` < 12000 ~ "Middle income",
    TRUE                     ~ "High income"
  ))|>
  filter(`Access to clean fuels and technologies for cooking` <= 100)  # remove bad values
```

```{r}
model1 <- lm(`Access to clean fuels and technologies for cooking` ~ 
              log(`GDP per capita`), 
            data = data_2023)
summary(model1)
```

```{r}
library(tidyverse)
data_2023 <- data_2023 |>
  mutate(log_gdp = log(`GDP per capita`))

ggplot(data_2023, aes(x = log_gdp,
                      y = `Access to clean fuels and technologies for cooking`)) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "lm",
              formula = y ~ x + I(x^2)) +
  labs(x = "Log GDP per capita",
       y = "Access to clean cooking fuels (%)")
```
```{r}
model2 <- lm(`Access to clean fuels and technologies for cooking` ~ log(`GDP per capita`) + I(log(`GDP per capita`)^2),
             data = data_2023)
summary(model2)
```
```{r}
anova(model1, model2)
```
```{r}
library(readr)
share_of_rural_population_with_electricity_access_vs_share_of_total_population_with_electricity_access <- read_csv("share-of-rural-population-with-electricity-access-vs-share-of-total-population-with-electricity-access.csv")
```
```{r}
rural_vs_urban <- share_of_rural_population_with_electricity_access_vs_share_of_total_population_with_electricity_access
```

```{r}
rural_vs_urban_2023 <- rural_vs_urban |>
  filter(Year == 2023)
```
```{r}
rural_vs_urban_2023 <- as.data.frame(rural_vs_urban_2023)
```
```{r}
joined_data <- full_join(data_2023, rural_vs_urban_2023, by = "Entity")
```
```{r}
joined_data <- joined_data |>
  mutate(rural_urban_gap = `Share of urban population with access` - 
          `Share of rural population with access`)
```
```{r}
model <- lm (`Access to clean fuels and technologies for cooking`~ log_gdp + rural_urban_gap, data = joined_data)
summary(model)
```
```{r}

```

## Results 
1. When fitting a linear relationship between GDP per capita and access to clean cooking fuels, a one-unit increase in log GDP per capita is associated with an average increase of 26.98% in access to clean cooking fuels. Very statistically significant. 

2. [Maybe remove this component? Not sure if needed] We noticed that the linearity assumption may not be fulfilled when fitting a linear relationship, so we tried fitting a quadratic relationship between GDP and access, which appeared to be a better fit, suggesting that for low-income countries, rise in GDP is important for increase fuel access, but this plateaus at a certain GDP level. ANOVA test affirms that the quadratic is a better fit. 

3. We then added a control for difference in % of urban and rural population with access to clean cooking fuels in a given country. While controlling for this rural/urban gap, a one-unit increase in log GDP is associated with an average increase of 19.50212% in access to clean cooking fuels. When holding log GDP constant, a one-unit increase in the gap in access between rural and urban population in a country is associated with a 0.56202% decrease in access to clean cooking fuels (statistically significant), suggesting that equality within a country, despite GDP, is also important for access. 

4. As a final component to make more public health related -- do a regression for access to clean cooking fuels and deaths from indoor pollution. 

## Discussion 